### 2021-07-05 rad-icfp-2021/unf.tex
simple writing phase
writing a bit in correctness
Few fixes (+ Jacobian defined as a math operator)

\section{UNF}

@@

There are a few notable things in this syntax. 
Every term is indexed by a well-defined contex $\Gamma$ from the source language.
In addition, a variable has a integer index $1\leq i\leq n$ where $n$ is the size of the context $\Gamma$.
Every constant, unary, or binary operator is summarised as an $n$-ary operator !op!$_n$.
Sequential composition is denoted by !;! and !e1;e2! means that !e1! should be performed, and then !e2!.
Every array operator !map!, !map2! and !foldl! has an extra index which represents a well-formed term in the source language.

The types are the same as in the source language. 
A context $\Gamma=\{$!x1: T1,...,xn:Tn!$\}$ can always be seen as a type !T1!$\times$...$\times$!Tn!.
UNF is very syntactic and does not really use variables. 
In fact it consists only of constants and composition. 
For this reason, we denote contexts as types in the indices of the terms, and $\times$ is used for context extension.
But we do emphasize that we mean contexts and the actual variables are remembered and
this will be key when compiling to the target language after differentiation.

@@

@@

Now our program in UNF is in an equivalent form to straight-line programs. 
Before presenting our differentiation macro, we present the target language for this macro. 
We call this intermediate representation target UNF.
The main idea is similar to source UNF where we only have constants and sequential composition.
That said, the key difference is that now every constant represents a pair of an operation and what is meant to represent its Jacobian.
Following the insights from Section~\ref{sec:background}, the composition of pairs reverses the order on the second component. T
his represents the fact that reverse-mode reverts the computation flow, 
and that pre-composition (using a continuation) is the simplest way to do this in a purely functional setting.
On pairs, !;! still represents sequential composition for the first component, and reversed composition is represented by $\circ$ on the second component.
The grammar and types are given in Figure~\ref{fig:unf_target_grammar}. 
For every constant $C$ from the source UNF, we have a new constant $J^TC$ representing its transpose Jacobian.
One thing to note is that one can only form pairs in this language. 
Similarly to the source UNF, sequential composition is the only operation which is not a constant of the language.

@@

Technically, all our terms in target UNF should have an extra index $\rho$ which is a type from Source.
This will be the return type of the continuation for the jacobian accumulation. 
In other words, if we want the full gradient of a term $\Gamma\vdash$!e!: $\reals$, $\rho$ should be chosen to be $\Gamma$.
For every type !T!, let $\Dtype$(!T!)$:=$!T!$\times$(!T!->$\rho$). 
The typesystem is given in Figure~\ref{fig:target_unf_typesystem}.

@@

% subsection transformations to and from UNF (end)

>>>>

\section{UNF}
\label{sec:unf}

@@

There are a few notable things in this syntax. 
Every term is indexed by a well-defined context $\Gamma$ from the source language.
In addition, a variable has a integer index $1\leq i\leq n$ where $n$ is the size of the context $\Gamma$.
Every constant, unary, or binary operator is summarized as an $n$-ary operator !op$_n$!.
Sequential composition is denoted by !;! and !e$_{1}$;e$_{2}$! means that !e$_{1}$! should be performed, and then !e$_{2}$!.
Every array operator !map!, !map2! and !foldl! has an extra index which represents a well-formed term in the source language.

The types are the same as in the source language. 
A context $\Gamma=\{$!x$_{1}$: T$_{1}$,...,x$_n$:T$_n$!$\}$ can always be seen as a type !T$_{1}$$\times$...$\times$ T$_n$!.
UNF is very syntactic and does not really use variables. 
In fact it consists only of constants and composition. 
For this reason, we denote contexts as types in the indices of the terms, and $\times$ is used for context extension.
But we do emphasize that we mean contexts and the actual variables are remembered and
this will be key when compiling to the target language after differentiation.

@@

\begin{example}
    TODO: simple term in the language, explain intuition.
\end{example}

@@

Now our program in UNF is in an equivalent form to straight-line programs. 
Before presenting our differentiation macro, we present the target language for this macro. 
We call this intermediate representation target UNF.
The main idea is similar to source UNF where we only have constants and sequential composition.
That said, the key difference is that now every constant represents a pair of an operation and what is meant to represent its Jacobian.
Following the insights from Section~\ref{sec:background}, the composition of pairs reverses the order on the second component. T
his represents the fact that reverse-mode reverts the computation flow, 
and that pre-composition (using a continuation) is the simplest way to do this in a purely functional setting.
On pairs, !;! still represents sequential composition for the first component, and reversed composition is represented by $\circ$ on the second component.
The grammar and types are given in Figure~\ref{fig:unf_target_grammar}. 
For every constant $C$ from the source UNF, we have a new constant $\J^TC$ representing its transpose Jacobian.
One thing to note is that one can only form pairs in this language. 
Similarly to the source UNF, sequential composition is the only operation which is not a constant of the language.

@@

Technically, all our terms in target UNF should have an extra index $\rho$ which is a type from Source.
This will be the return type of the continuation for the Jacobian accumulation. 
In other words, if we want the full gradient of a term $\Gamma\vdash$!e:$\reals$!, $\rho$ should be chosen to be $\Gamma$.
For every type !T!, let !$\Dtype($!T!$):= \,$T$\times$(T->$\rho$)!. 
The type system is given in Figure~\ref{fig:target_unf_typesystem}.

@@

The transformation from the target UNF to the target language is presented in Figure~\ref{fig:unf_to_target}.

\input{figures/unf_to_target}
